\documentclass[
  pagePreset=largepage,
  babelLanguage=british,
]{aruno-anecdote}

\title{Forest Path}

\usepackage{mylayout}

\begin{document}

%\input{./src-tex/titlepage.tex}
%
%\clearpage
%\input{./src-tex/copyright.tex}
%
%\clearpage
%\input{./src-tex/frontmatter.tex}
%
%\clearpage
%\tableofcontents
%
%\chapter{Preface to Second Edition, 2012}
%\input{./src-tex/preface-to-second-edition-2012.tex}
%
%\chapter{Note}
%Please be aware that the authors' monastic titles when the
%original essays were written have been kept (e.g. `Sāmaṇera' or `Tan').
%You will see in the short summaries added after each essay, which
%explain what has happened since then, that almost all of them are now
%senior monks whom we would usually honour by calling them `Ajahns',
%teachers.
%
%\chapter{Luang Por}
%\input{./src-tex/luang-por.tex}

\setChapterAuthor{Ajahn Jayasāro}
\setChapterNote{An excerpt from the biography of Ajahn Chah dealing with the arrival of Ajahn Sumedho at Wat Pah Pong.}
\chapter{Twain Shall Meet}
\markright{\chapterAuthor}
\input{./src-tex/twain-shall-meet.tex}

\setChapterAuthor{Ajahn Jayasāro}
\setChapterNote{An excerpt from the biography of Ajahn Chah containing advice for meditators.}
\chapter{Doubt and Other Demons}
\markright{\chapterAuthor}
\input{./src-tex/doubt-and-other-demons.tex}

\setChapterAuthor{Ajahn Pasanno}
\setChapterNote{An edited version of a Dhamma talk given at Wat Pah Nanachat on 18 November 1995.}
\chapter{Keep it Simple}
\markright{\chapterAuthor}
\input{./src-tex/keep-it-simple.tex}

\setChapterAuthor{Ajahn Vipassi}
\chapter[Facets of Life at Wat Pah Nanachat]{Facets of Life at\newline Wat Pah Nanachat}
\markright{\chapterAuthor}
\input{./src-tex/facets-of-life.tex}

\setChapterAuthor{Ajahn Jayasāro}
\setChapterNote{An edited version of a Dhamma talk given to the Sangha during the 1998 Rains Retreat.}
\chapter{The Beauty of Sīla}
\markright{\chapterAuthor}
\input{./src-tex/the-beauty-of-sila.tex}

\setChapterAuthor{Tan Saññamo}
\setChapterNote{A young monk's reflections on working with discomfort.}
\chapter{Mindfulness with Mosquitoes}
\markright{\chapterAuthor}
\input{./src-tex/mindfulness-with-mosquitoes.tex}

\setChapterAuthor{Sāmaṇera Khemavāro}
\setChapterNote{A novice considers the changes in his attitudes to life following ordination.}
\chapter{Progress without Movement}
\markright{\chapterAuthor}
\input{./src-tex/progress-without-movement.tex}

\setChapterNote{An interview with three Thai monks practising at Wat Pah Nanachat.}
\chapter{No Thai, No Farang}
\markright{\chapterAuthor}
\input{./src-tex/no-thai-no-farang.tex}

% Life and Times at Wat Pah Nanachat -- The Teaching of My Life
\setChapterAuthor{Christine Lem}
\setChapterNote{A lay meditator considers the challenges and benefits of practising at Wat Pah Nanachat.}
\chapter[Ten Thousand Joys and Ten Thousand Sorrows]{Ten Thousand Joys and\newline Ten Thousand Sorrows}
\markright{\chapterAuthor}
\input{./src-tex/ten-thousand.tex}

\setChapterAuthor{Sāmaṇera Gunavuddho}
\setChapterNote{Exploring identity and the similarities between his training in music and monastic life.}
\chapter{Hey, man, don't give up your music!}
\markright{\chapterAuthor}
\input{./src-tex/hey-man.tex}

\setChapterNote{Two local lay supporters talk about their relationship to the wat.}
\chapter{Refuge in the Sangha}
\markright{\chapterAuthor}
\input{./src-tex/refuge-in-sangha.tex}

\setChapterAuthor{Ajahn Jayasāro}
\setChapterNote{An interview conducted by Tan Paññāvuddho.}
\chapter{Dtao Dam Forest Monastery}
\markright{\chapterAuthor}
\input{./src-tex/dtao-dam.tex}

\setChapterAuthor{Paññāvuddho Bhikkhu}
\setChapterNote{Practising in the ancient way at the foot of a tree.}
\chapter{Learning Forest Dhamma}
\markright{\chapterAuthor}
\input{./src-tex/learning-forest-dhamma.tex}

\setChapterAuthor{Ajahn Piak}
\setChapterNote{An interview with a senior disciple of Ajahn Chah conducted by\linebreak Ajahn Chandako.}
\chapter{What it Takes to Reach the Goal}
\markright{\chapterAuthor}
\input{./src-tex/what-it-takes.tex}

\setChapterAuthor{Tan Acalo}
\setChapterNote{A monk visits his family after three years at Wat Pah Nanachat.}
\chapter{Acceptance, Forgiveness and Deep Blue Empathy\ldots{} Going Home}
\markright{\chapterAuthor}
\input{./src-tex/acceptance-forgiveness.tex}

\setChapterAuthor{Tan Yātiko}
\setChapterNote{An account of a period practising at the monastery of Tan Ajahn Mahā Boowa.}
\chapter{Two months at Wat Pah Baan Taad}
\markright{\chapterAuthor}
\input{./src-tex/two-months.tex}

\setChapterAuthor{Ajahn Jayasāro}
\setChapterNote{An edited version of a Dhamma talk given to the Sangha during the Rains Retreat of 1998.}
\chapter{Faith in the Quest}
\markright{\chapterAuthor}
\input{./src-tex/faith-in-the-quest.tex}

%\chapter{Coming to Visit Wat Pah Nanachat}
%\input{./src-tex/coming-to-visit.tex}
%
%\chapter{Transliteration of Thai Words}
%
%\chapter{Glossary}

\end{document}
